---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r opts, echo = FALSE}
  knitr::opts_chunk$set(
     fig.path = "images/"
  )
```

## Loading and preprocessing the data

First at all loads the required libraries

```{r}
library(dplyr, warn.conflicts=F, quietly=T)
library(tidyr, warn.conflicts=F, quietly=T)
library(ggplot2, warn.conflicts=F, quietly=T)
library(gridExtra, warn.conflicts=F, quietly=T)
```

Data is stored into data subdirectory.
Usually, the first step should be take a look into raw data from command line: In example with more command. But in this case wew know is a csv file

We try to read the file as csv, then we take a look to data frame with head (beginning of file) and tail (end of file) and structure

```{r}
df <- read.csv("../data/repdata_data_StormData.csv.bz2", na.strings="")
summary(df)
```

## Question 1: Across the United States, which types of events are most harmful with respect to population health?

According the info provided into document: "*NATIONAL WEATHER SERVICE INSTRUCTION 10-1605*" Impact on population health, as direct as indirect is stored into field **INJURIES**. So take a look at Injuries by type of Event


```{r}
plot(INJURIES ~ EVTYPE, data=df, main="Injuries by Event Type")
```

### Analyzing outlier

The event type with the second higher value looks a bit strange, so we dive into:

1.- First, order the data frame by Injuries in descending order

2.- Retrieve the second record to identify the event type

3.- Analyze this event 

```{r}
dfs <- df[order(-df$INJURIES),]
dfs[2,"EVTYPE"]
dfs <- subset(dfs, EVTYPE == dfs[2,"EVTYPE"] )
summary(dfs$INJURIES)
```

As we can see, even though the max is 1,568 the mean is almost one, so we can consider this event as an exceptional situation.


### Comparing frequencies against quantities

In order to answer to question we need to consider two situations:

A. Events not very usuals but highly dangerous
B. Frequent events although not very dangerous

Lets compare these situtations getting the top five by mean and and total:

```{r}
par(mfrow=c(1,2))
grouped <- df %>% group_by(EVTYPE) %>% summarise(Total=n(), Mean=mean(INJURIES))
dfMean <- grouped[order(-grouped$Mean),]
dfTotal<- grouped[order(-grouped$Total),]
grid.arrange(tableGrob(dfMean[1:5,]),
             tableGrob(dfTotal[1:5,]),
             nrow=1)
```

On the left we can see not very frequency events but dangerous; in example,  we have only 4 Wild Fires but each time this time occurs a lot of people is injuried.

On the right side we have got another vision, each event is not very harmful but they are so frequents that we need to consider as dagerous for population.

So, in my opinion, the answer to question is:

** 1. Tornados** They are very common
** 2. Tropical Storms and Winds** Also common in South of Country and eventually very harmful
** 3. Heat Wave**
** 4. Wild Fires** It looks exists protocols against this because there are not many events, but the mean of injuries is very high: 37.5


## Question 2: Across the United States, which types of events have the greatest economic consequences?

According the info provided into document: "*NATIONAL WEATHER SERVICE INSTRUCTION 10-1605*" Impact on economy is stored into **PROPDMG** and **PROPDMGEXP**. this one indicates the units of previous one.

### Normalize data

First step is to use the same unit for all damages, the problem is PROPDMGEXP is not normalized:
```{r}
unique(df$PROPDMGEXP)
```

So we asume these criteria:

1.- N/A = 1 (Normally N/A is associated with 0 value)
2.- Number is a multiplier
3.- B/M/K/T/H as usual: Billion, Million, ....
4.- Other data assume 1 unit

```{r}
par(mfrow=c(1,1))
calculate <- function(Amount, Unit) {
      if (is.na(Unit)) return (Amount * 1)
      if (is.numeric(Unit)) return (Amount * Unit)
      if (Unit == "B" || Unit == "b") return (Amount * 1000000000)
      if (Unit == "M" || Unit == "M") return (Amount * 1000000)
      if (Unit == "K" || Unit == "k") return (Amount * 1000)
      if (Unit == "H" || Unit == "h") return (Amount * 100)
      Amount
}
dft <- df %>% rowwise() %>% mutate(Cost = calculate(PROPDMG, PROPDMGEXP))
```

And now take a look on data:

```{r}
summary(dft)
```

Again, there are some "*outliers*", because mean of Cost is 142 USD but there are at least one which cost is more than 91,000.

However, when we get the most expensive events:

```{r}
Sorted <- dft[order(-dft$Amount),]
head(Sorted)
```


We can see that the event types matchs the previous conclusions related to human injuries. So the answer to question can be the same

### Analyzing outlier

The event type with the second higher value looks a bit strange, so we dive into:

1.- First, order the data frame by Injuries in descending order

2.- Retrieve the second record to identify the event type

3.- Analyze this event 

```{r}
dfs <- df[order(-df$INJURIES),]
dfs[2,"EVTYPE"]
dfs <- subset(dfs, EVTYPE == dfs[2,"EVTYPE"] )
summary(dfs$INJURIES)
```

As we can see, even though the max is 1,568 the mean is almost one, so we can consider this event as an exceptional situation.


### Comparing frequencies against quantities

In order to answer to question we need to consider two situations:

A. Events not very usuals but highly dangerous
B. Frequent events although not very dangerous

Lets compare these situtations getting the top five by mean and and total:

```{r}
par(mfrow=c(1,2))
grouped <- df %>% group_by(EVTYPE) %>% summarise(Total=n(), Mean=mean(INJURIES))
dfMean <- grouped[order(-grouped$Mean),]
dfTotal<- grouped[order(-grouped$Total),]
grid.arrange(tableGrob(dfMean[1:5,]),
             tableGrob(dfTotal[1:5,]),
             nrow=1)
```

On the left we can see not very frequency events but dangerous; in example,  we have only 4 Wild Fires but each time this time occurs a lot of people is injuried.

On the right side we have got another vision, each event is not very harmful but they are so frequents that we need to consider as dagerous for population.

So, in my opinion, the answer to question is:

** 1. Tornados** They are very common
** 2. Tropical Storms and Winds** Also common in South of Country and eventually very harmful
** 3. Heat Wave**
** 4. Wild Fires** It looks exists protocols against this because there are not many events, but the mean of injuries is very high: 37.5

